Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-09-21 analysis/paper/paper.Rmd
reducing word count for resubmission

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
# prepare for Antiquity https://antiquity.ac.uk/submit-paper
# should be no more than 13.5cm wide and at a minimum resolution of 300dpi. 
# not exceed 6000 words excluding figures, tables, highlights 
image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Studying the consumption of foreign goods can improve our understanding of the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Rubertone2000; @Scaramelli2005; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005].

Prior to contact with Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since the 10th century. Taiwan encountered large waves of Han Chinese immigration after 17th century when Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007]. The exchange network in northern Taiwan was influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on shipping routes [@Chen2005]. This intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007].

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonisation of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonisation. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is a physically isolated area adjacent to colonial centres and there are no traces of European settlement but occasional interactions with the Europeans. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements living beyond the borders of direct colonial occupation.

Although mountains isolated Yilan by land, rivers and seas provided a way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan, and Han Chinese. The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis before European were present. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. Trading activities brought a wide variety of objects into indigenous communities in Yilan. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore archaeological evidence from Yilan spanning from the late Iron Age to the historical period to address the question of whether indirect colonial influences on the indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf. @Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine the potential of ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European arrival due to participation in a larger scale exchange network. I also predict that after the European presence became established there will be new patterns in the spatial distributions of ornaments at Kiwulan that indicate increasing social inequality. I hypothesize that after 1662 when the European presence declined and the Chinese became the dominant international tradition partner, there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European arrival, and a lesser diversity after the Chinese presence became established, I compare frequencies of ornament types between different time periods. To test the prediction that after European arrival were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. An indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige  [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during traveling [@Kang2016, ch. 6]. Thus, I assume that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

# Excavations at Kiwulan in northeastern Taiwan
```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700 - 1200 AD) and a Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age to the historical period starting from the 17th century with the European presence in Taiwan.

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period (L5 and L6), European period (L4), and Chinese period (L2 and L1) to indicate the dominant foreign influence and cultural interaction for different time periods.

I identify the first European arrival at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. In the archaeological record, the approximate indicators of the start of the European influence at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

European colonisation ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial context are not analyzed here since not many burials can be assigned to a time period with confidence. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

The results show there was a greater diversity of ornaments types and materials at Kiwulan during the European presence. This might result from the large scale exchange network that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the mortuary practices of Han Chinese. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. According to Chinese census reports in 1821 and Japanese field surveys in 1895, the Kiwulan population was less than 100 [@Yao1996; @Ino1898].

>>>>

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  cache = TRUE,
  fig.path = "../figures/"
)
# prepare for Antiquity https://antiquity.ac.uk/submit-paper
# should be no more than 13.5cm wide and at a minimum resolution of 300dpi. 
# not exceed 6000 words excluding figures, tables, highlights 
image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in Taiwan, the northeast region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes ornaments excavated from Kiwulan (700-1900 AD), the largest Iron Age settlement on the Yilan plain in northeast Taiwan. The first recorded European presence in Yilan was a Spanish revenge attack on indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rules and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-2, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages.

@@

The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged agate beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of Indigenous people to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the members of the Kavalan tribe, the contemporary indigenous people of the Yilan plain, describes how agate beads were used in divination practices by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families. When not in use, beads were often hidden in safe places, such as on the roof beams. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. However, compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

# Excavations at Kiwulan in northeast Taiwan

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

Kiwulan (Figure \@ref(fig:KWL-map)) was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700-1200 AD) and a Upper Layer Culture (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the Upper Layer Culture because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

The European colonial presence in northern Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains, clay stoves. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares were excavated _in situ_ from household area of Kiwulan, indicated by numerous post-holes. Thirty burials contained 3173 ornaments, and 27 ornaments were recovered from middens. In the burials ornaments were in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial contexts are not analysed here since few burials can be confidently assigned to a time period. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

The results show there was a greater diversity of ornaments types and materials at Kiwulan during the European presence. This might result from the large scale exchange network that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the mortuary practices of Han Chinese. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].
